Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection)

### 2017-07-05 code/Rmd/ALL_REPROCESSED.Rmd
Minor modifications

### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Diagnosis", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

all.diff.exp = c(all.diff.exp, list(Diagnosis = DE))
all.fit = c(all.fit, list(Diagnosis = FIT))
```

@@

### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
Differential expression is performed on the BrainRegion.Diagnosis x Gender variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=15, fig.width=13, eval = F}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$BrainRegion.Diagnosis.Gender = factor(paste(COVARIATES$BrainRegion.Diagnosis,
                                                       COVARIATES$Gender,
                                                       sep = '_'))

# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(adjust.covars, 'Gender')), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

@@

# Fit contrast
# Fit contrast
cntr = c()
for (region in c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  tmp = combn(paste0('BrainRegion.Diagnosis.Gender', region, '_', c('AD', 'CONTROL')),2)
  tmp1 = combn(paste0())
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

cntr = c('BrainRegion.Diagnosis.SEXIFG.AD.MALE-BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXIFG.AD.FEMALE-BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXIFG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXIFG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXIFG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXFP.AD.MALE-BrainRegion.Diagnosis.SEXFP.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXFP.AD.FEMALE-BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE+0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXFP.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXFP.AD.MALE+0.5*BrainRegion.Diagnosis.SEXFP.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXPHG.AD.MALE-BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXPHG.AD.FEMALE-BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXPHG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXPHG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXPHG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXSTG.AD.MALE-BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE',
         'BrainRegion.Diagnosis.SEXSTG.AD.FEMALE-BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE+0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE',
         '0.5*BrainRegion.Diagnosis.SEXSTG.AD.FEMALE-0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.FEMALE
         -0.5*BrainRegion.Diagnosis.SEXSTG.AD.MALE+0.5*BrainRegion.Diagnosis.SEXSTG.CONTROL.MALE')
contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
colnames(contrast) = c('IFG.AD.MALE-IFG.CONTROL.MALE',
                       'IFG.AD.FEMALE-IFG.CONTROL.FEMALE',
                       'IFG.AD.ALL-IFG.CONTROL.ALL',
                       'IFG.ALL.FEMALE-IFG.ALL.MALE',
                       'IFG.ALL.ALL-IFG.ALL.ALL',
                       'FP.AD.MALE-FP.CONTROL.MALE',
                       'FP.AD.FEMALE-FP.CONTROL.FEMALE',
                       'FP.AD.ALL-FP.CONTROL.ALL',
                       'FP.ALL.FEMALE-FP.ALL.MALE',
                       'FP.ALL.ALL-FP.ALL.ALL',
                       'PHG.AD.MALE-PHG.CONTROL.MALE',
                       'PHG.AD.FEMALE-PHG.CONTROL.FEMALE',
                       'PHG.AD.ALL-PHG.CONTROL.ALL',
                       'PHG.ALL.FEMALE-PHG.ALL.MALE',
                       'PHG.ALL.ALL-PHG.ALL.ALL',
                       'STG.AD.MALE-STG.CONTROL.MALE',
                       'STG.AD.FEMALE-STG.CONTROL.FEMALE',
                       'STG.AD.ALL-STG.CONTROL.ALL',
                       'STG.ALL.FEMALE-STG.ALL.MALE',
                       'STG.ALL.ALL-STG.ALL.ALL')

@@

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MSSM',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=15, fig.width=13, eval = F}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], setdiff(postAdjustCovars, 'AOD')), drop = F], Intercept = F)
ind = grep('Diagnosis', colnames(DESIGN$design))
DESIGN$design[,ind] = DESIGN$design[,ind] * COVARIATES[rownames(DESIGN$design),'AOD']
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
colnames(DESIGN$design)[ind] = gsub('BrainRegion.Diagnosis', '', paste0(colnames(DESIGN$design)[ind], '.AOD'))

@@

# Fit contrast
cntr = c("FP.AD.AOD-FP.CONTROL.AOD",
         "0.5*FP.AD.AOD+0.5*FP.CONTROL.AOD",
         "IFG.AD.AOD-IFG.CONTROL.AOD",
         "0.5*IFG.AD.AOD+0.5*IFG.CONTROL.AOD",
         "PHG.AD.AOD-PHG.CONTROL.AOD",
         "0.5*PHG.AD.AOD+0.5*PHG.CONTROL.AOD",
         "STG.AD.AOD-STG.CONTROL.AOD",
         "0.5*STG.AD.AOD+0.5*STG.CONTROL.AOD")
contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
colnames(contrast) = c("FP.AD.AOD-FP.CONTROL.AOD",
                       "FP.ALL.AOD",
                       "IFG.AD.AOD-IFG.CONTROL.AOD",
                       "IFG.ALL.AOD",
                       "PHG.AD.AOD-PHG.CONTROL.AOD",
                       "PHG.ALL.AOD",
                       "STG.AD.AOD-STG.CONTROL.AOD",
                       "STG.ALL.AOD")
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MSSM',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

>>>>

```{r decompse.RESIDUALS, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

### Residual calculation
Calculate weighted residuals and add back "BrainRegion.Diagnosis" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Diagnosis", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

all.diff.exp = list(Diagnosis = DE)
all.fit = list(Diagnosis = FIT)
```

@@

### Differential expression analysis (with BrainRegion.Diagnosis and Gender)
Differential expression is performed on the BrainRegion.Diagnosis x Gender variable by controlling for covariates identified above
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=7, fig.width=24}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$BrainRegion.Diagnosis.Gender = factor(paste(COVARIATES$BrainRegion.Diagnosis,
                                                       COVARIATES$Gender,
                                                       sep = '_'))

# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('BrainRegion.Diagnosis.Gender', setdiff(adjust.covars, c('BrainRegion.Diagnosis', 'Gender'))), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

@@

# Fit contrast
cntr = c()
for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  cnt = c(paste0('BrainRegion.Diagnosis.Gender', region, '_AD_MALE-BrainRegion.Diagnosis.Gender', region, '_CONTROL_MALE'),
          paste0('BrainRegion.Diagnosis.Gender', region, '_AD_FEMALE-BrainRegion.Diagnosis.Gender', region, '_CONTROL_FEMALE'),
          paste0('0.5*BrainRegion.Diagnosis.Gender', region, '_AD_MALE+0.5*BrainRegion.Diagnosis.Gender', region, 
                 '_AD_FEMALE-0.5*BrainRegion.Diagnosis.Gender', region, '_CONTROL_MALE-0.5*BrainRegion.Diagnosis.Gender',region, '_CONTROL_FEMALE'),
          paste0('0.5*BrainRegion.Diagnosis.Gender', region, '_AD_FEMALE+0.5*BrainRegion.Diagnosis.Gender', region, 
                 '_CONTROL_FEMALE-0.5*BrainRegion.Diagnosis.Gender', region, '_AD_MALE-0.5*BrainRegion.Diagnosis.Gender',region, '_CONTROL_MALE'),
          paste0('0.5*BrainRegion.Diagnosis.Gender', region, '_AD_FEMALE-0.5*BrainRegion.Diagnosis.Gender', region, 
                 '_CONTROL_FEMALE-0.5*BrainRegion.Diagnosis.Gender', region, '_AD_MALE+0.5*BrainRegion.Diagnosis.Gender',region, '_CONTROL_MALE'))
  cntr = c(cntr, cnt)
}
contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
cnames = c()
for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  cnt = c(paste0(region, '_AD_MALE-', region, '_CONTROL_MALE'),
          paste0(region, '_AD_FEMALE-', region, '_CONTROL_FEMALE'),
          paste0(region, '_AD_ALL-', region, '_CONTROL_ALL'),
          paste0(region, '_ALL_FEMALE-', region, '_ALL_MALE'),
          paste0(region, '_ALL_ALL-', region,'_ALL_ALL'))
  cnames = c(cnames, cnt)
}
colnames(contrast) = cnames

@@

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('Gene.ID')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(Gene.ID, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(Gene.ID, Comparison, Direction) %>%
  group_by(Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Gene.ID))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=7, fig.width=24}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(setdiff(adjust.covars, 'AOD')), drop = F], Intercept = F)
ind = grep('BrainRegion.Diagnosis', colnames(DESIGN$design))
DESIGN$design[,ind] = DESIGN$design[,ind]*COVARIATES$AOD
colnames(DESIGN$design)[ind] = paste(colnames(DESIGN$design)[ind], 'AOD', sep = '_')
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

@@

# Fit contrast
cntr = c()
for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  cnt = c(paste0('BrainRegion.Diagnosis', region, '_AD_AOD-BrainRegion.Diagnosis', region, '_CONTROL_AOD'),
          paste0('0.5*BrainRegion.Diagnosis', region, '_AD_AOD+0.5*BrainRegion.Diagnosis', region, '_CONTROL_AOD'))
  cntr = c(cntr, cnt)
}
contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
cnames = c()
for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  cnt = c(paste0(region, '_AD_AOD-', region, '_CONTROL_AOD'),
          paste0(region, '_ALL_AOD'))
  cnames = c(cnames, cnt)
}
colnames(contrast) = cnames

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('Gene.ID')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(Gene.ID, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(Gene.ID, Comparison, Direction) %>%
  group_by(Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(Gene.ID))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

@@

### Differential expression analysis (with BrainRegion.Diagnosis and Celltype Fractions)
Differential expression is performed on the BrainRegion.Diagnosis x CelltypeFractions variable by controlling for covariates identified above

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=7, fig.width=24, eval = F}
```
